---
title: "Inferring ages over 90 in GENIE BPC"
subtitle: "Reidentification attacks - now with arithmetic"
format: pdf
mainfont: Fira Sans
execute: 
  echo: false
  include: false
---



```{r}
#| include: false
library(purrr); library(here); library(fs)
purrr::walk(.x = fs::dir_ls(here('R')), .f = source)

max_inferred_ages_pt <- readr::read_rds(
  here('data', 'max_inferred_ages_pt.rds')
)
```


## Introduction

In GENIE we hope to follow the HIPAA safe harbor method on age, which reads:

> All elements of dates (except year) for dates that are directly related to an individual, including birth date, admission date, discharge date, death date, and all ages over 89 and all elements of dates (including year) **indicative of such age**, except that such ages and elements may be aggregated into a single category of age 90 or older.

We recently completed an assessment of **directly stated** intervals in GENIE which would reveal a patient's age as over 89.  For example, the interval from birth to diagnosis, or the interval from diagnosis to death (however unlikely it may be that interval is over 89 years).  However, "indicative of such age" may go beyond this naive assessment (TBD).  It also seems reasonable to assess whether two fields combined could reveal an age over 89.

This document is an initial exploration into how many people can be revealed to be over 89 using no external data sources and elementary arithmetic operations alone.  That is, we might add together two intervals if they are ajoining, or subtract two calendar years from one another, but we won't be doing anything fancy like record linkage or imputation.  Therefore, we can't disregard this issue on the basis that it requires too much technical skill or time to do (and therefore poses a 'barrier').

Here we do not deal with the privacy question of whether or not it **is** a problem to be able to reconstruct ages from data elements.  A governance and legal group will need to provide an ethical guideline on that.  Our goal is only providing an assessment of the scale of ages over 89 which can be reconstructed this way in the GENIE data.


```{r}
dfp_violations <- max_inferred_ages_pt %>%
  group_by(cohort) %>%
  summarize(
    n_total = n(),
    
    n_gte_90 = sum(max_inferred_age >= (90 - 0.5/365.25), na.rm = T),
    gte_90 = n_pct_str(n_gte_90, n_total, show_d = T, digits = 3),
    
    # why plus here? 89 plus one day, minus one half day for rounding tolerance.
    n_gt_89 = sum(max_inferred_age >= (89 + 0.5/365.25), na.rm = T),
    gt_89 = n_pct_str(n_gt_89, n_total, show_d = T, digits = 3),
    
    .groups = "drop"
  ) %>%
  select(cohort, gte_90, gt_89)
```


## Results

The following table lists the number of people in two categories:

1. Greater than or equal to 90 years old (`gte_90`).  Clear issues.
2. Greater than or equal to 89 years plus one day old (`gt_89`).  This is a superset of the previous case, of course.  These are issues if we use the pervasive "89 plus one day" interpretation of the law.  As we've previously pointed out, this interpretation is hard to support if one reads to the end of the quoted HIPAA sentence, but it retains relevance due to inclusion in the GENIE SOPs for now.
 
```{r}
#| include: true 
dfp_violations %>%
  flextable(.) %>%
  autofit(.)
```


**Observations:**

- The cohorts which have a problem appear to be the same two that had problems with directly stating ages over 89 for `age_dx`.
- Compared to the previous problems noticed with `age_dx`, the numbers are higher here.  Therefore we can tell there are additional potential issues here (without the need to cross reference).










## Methods

We used the most recent consortium release for each cohort.

Our method for ages consists of adding two intervals together for each case:  (1) the maximum interval from anything to diagnosis, and (2) the maximum interval from diagnosis to any other observation.  The reason we did this is that diagnosis is by far the most common start time for intervals in GENIE BPC, and (almost) everyone has a birth to diagnosis interval. Therefore, we can cover most of the ground that exists with minimal effort by doing this.  Before diagnosis we only considered the various names for birth to diagnosis.  For the observations after diagnosis we included:

- Death or censoring time for OS.
- Sequencing.
- Distant metastasis times.
- Regimen stop (or last known administration) times.

Pathology, imaging and med onc times were not included due to having a different anchor (start) point.  They index from the first BPC cancer (rather than a specific BPC cancer) and our joins mostly rely on `{record_id, ca_seq}`.  These could be included later, but it didn't seem worthwhile for this analysis.  Notably, if the dataset is constructed correctly, `tt_os_dx_*` should be the maximum interval from diagnosis to any observation, so even including other variables is getting into the territory of QC.

To be a bit more specific, The following table states the variables which were found to contain the maximum interval **TO** diagnosis.  The counts show people:

```{r}
#| include: true
max_inferred_ages_pt %>%
  count(var_to_dx, sort = T) %>%
  head(5) %>%
  flextable(.) %>%
  autofit(.)
```
Note that those are all the same variable semantically with different names and rounding conventions.  Here is the same table for the variables **FROM** diagnosis:

```{r}
#| include: true
max_inferred_ages_pt %>%
  count(var_from_dx, sort = T) %>%
  head(5) %>%
  flextable(.) %>%
  autofit(.)
```

The `_mos` version of these variables tends to be greater than the others due to rounding when converting back to years.  This is not meaningful, on the order of hundredths of a day.  

A rounding tolerance was applied to the calculations.  For example, if we calculated that someone's age was 89.9999, that counts as 90.  The tolerance size is one-half day, but it probably isn't consequential.




## Future expansions

The most important way to expand this work is to **talk with a governance or legal team** about whether this situation is an issue.  Some analytical options for expansion after that:

- Consider differences between calendar years.  For example, what is the minimum distance implied by taking the difference of birth year and sequencing year?  We have not done this yet.
- Expand to other releases.  This requires a review of variable names due to the inconsistency of BPC naming over time.
- Expand to more variables (more a QC step).
- Expand to use other "middle" times (where cancer diagnosis X was our middle time here).  For example, there are intervals on both sides of CPT and metastasis which we could combine to reconstruct ages.
- External review of findings.


## Conclusions

If it is a problem to have the ability to reconstruct ages using the simple methods we've used here, then we have some additional problems here which require retraction/fixing/something.  The magnitude of the problem is smaller than anticipated.  Before doing additional work we should figure out if the issues presented here need to be corrected in GENIE from a governance or privacy perspective.



